---
title: "Main RMD"
author: "Brad Gannon"
date: "April 23, 2019"
output: html_document
---

```{r}
library(tidyverse)

knitr::opts_chunk$set(fig.width = 15, fig.height=10) # change figure size in Rmd/HTML

# Load data from CSV
data <- read.csv('MinimumWageData.csv')
names(data)

head(data, 2)

# Logical to delete Table_Data and Footnote columns, these are just notes and not useful

drops <- c("Table_Data", "Footnote")

data <- data[, !(names(data) %in% drops)]

head(data, 3)
```

```{r}
# See if data types in columns are logical
str(data)

```

The data was already cleaned on Kaggle, so the cleaning is very minimal. Outside of deleting the two columns that did not contain valuable information, the NA values, data types, factoring, etc. was already done.

```{r}
# For graphing, I will divide the data into pre-1990 and 1990-2017

pre1990 <- filter(data, Year < 1990)
post1990 <- filter(data, Year >= 1990)

# Breakdown of data from each year
p <- ggplot(pre1990, aes(x=Year)) + geom_histogram(color="black", fill="white", binwidth=1)
p <- p + xlab("Year") + ylab("Count of Data Points") + ggtitle("Data Points by Year (Pre-1990)")
p

p <- ggplot(post1990, aes(x=Year)) + geom_histogram(color="black", fill="gray", binwidth=1)
p <- p + xlab("Year") + ylab("Count of Data Points") + ggtitle("Data Points by Year (Post-1990)")
p

# These histograms show that the data for each year is all even, so the comparisons should be fairly accurate

# We will look at the spread of High.Value and Low.Value for each state by year

p <- ggplot(pre1990, aes(x=Year, y=High.Value, color=State)) + geom_boxplot(outlier.color="red", outlier.shape=8, outlier.size=2)
p <- p + xlab("Year") + ylab("Hourly Wage ($)") + ggtitle("Highest Minimum Wage Across US States pre-1990")
p <- p + facet_wrap(pre1990$State) + theme(legend.position="none")
p

p <- ggplot(pre1990, aes(x=Year, y=Low.Value, color=State)) + geom_boxplot(outlier.color="red", outlier.shape=8, outlier.size=2)
p <- p + xlab("Year") + ylab("Hourly Wage ($)") + ggtitle("Lowest Minimum Wage Across US States pre-1990")
p <- p + facet_wrap(pre1990$State) + theme(legend.position="none")
p

p <- ggplot(post1990, aes(x=Year, y=High.Value, color=State)) + geom_boxplot(outlier.color="red", outlier.shape=8, outlier.size=2)
p <- p + xlab("Year") + ylab("Hourly Wage ($)") + ggtitle("Highest Minimum Wage Across US States post-1990")
p <- p + facet_wrap(post1990$State) + theme(legend.position="none")
p

p <- ggplot(post1990, aes(x=Year, y=Low.Value, color=State)) + geom_boxplot(outlier.color="red", outlier.shape=8, outlier.size=2)
p <- p + xlab("Year") + ylab("Hourly Wage ($)") + ggtitle("Lowest Minimum Wage Across US States post-1990")
p <- p + facet_wrap(post1990$State) + theme(legend.position="none")
p
```

```{r}
# Look at trend of High.Value and Low.Value across the years overall
p <- ggplot(pre1990, aes(x=Year, y=High.Value)) + geom_violin(trim = FALSE)
p <- p + stat_summary(fun.y=median, geom="line", shape=18, size=2, color="red")
p <- p + ylab("High Minimum Wage ($)") + xlab("Year") + ggtitle("High Minimum Wage Trend pre-1990")
p

p <- ggplot(pre1990, aes(x=Year, y=Low.Value)) + geom_violin(trim = FALSE)
p <- p + stat_summary(fun.y=median, geom="line", shape=18, size=2, color="blue")
p <- p + ylab("Low Minimum Wage ($)") + xlab("Year") + ggtitle("Low Minimum Wage Trend pre-1990")
p

p <- ggplot(post1990, aes(x=Year, y=High.Value)) + geom_violin(trim = FALSE)
p <- p + stat_summary(fun.y=median, geom="line", shape=18, size=2, color="green")
p <- p + ylab("High Minimum Wage ($)") + xlab("Year") + ggtitle("High Minimum Wage Trend post-1990")
p

p <- ggplot(post1990, aes(x=Year, y=Low.Value)) + geom_violin(trim = FALSE)
p <- p + stat_summary(fun.y=median, geom="line", shape=18, size=2, color="purple")
p <- p + ylab("Low Minimum Wage ($)") + xlab("Year") + ggtitle("Low Minimum Wage Trend post-1990")
p
```

```{r}
# We should look at High.2018 and Low.2018 data in a similar way, but I left off here because it is almost midnight

```

All states over time -- ggplotly
```{r}
library(plotly)
p2 <- ggplot(data, aes(x=Year, y=High.Value, by=State)) + geom_line(aes(col=State)) 
ggplotly(p2)
```


How do states compare with the Federal wage? Which states closely follow it? Which states are above? Below?
```{r}
library(dplyr)
#create new column with difference from Federal of the same year
fed <- data %>% filter(State == 'Federal (FLSA)') %>% select(Year, High.Value)  #just Fed data
fed$fed.High.Value <- fed$High.Value #duplicates column with new name
fed <- fed %>% select(Year, fed.High.Value) #drops old name column (because I couldn't figure out an easier way)

eras <- fed %>% dplyr::group_by(fed.High.Value) %>% mutate(xmin=min(Year), xmax=max(Year)) #adds col for first and last year 

data2 <- data %>% group_by(State) %>% mutate(fed_diff = High.Value - fed$fed.High.Value)
newdata <- merge(x=data2, y=eras)
eralines <- geom_vline(xintercept = newdata$xmin, alpha=0.25)
eralabels <- geom_text(data=newdata, x=newdata$xmin, y=4, label=newdata$fed.High.Value, angle=90, size=3, hjust=0.2)
zeroline <- geom_hline(yintercept = 0)
plotlabs <- labs(title = "State Minimum Wage -- difference from Federal minimum wage",
                y = 'Difference ($)')
p3 <- ggplot(newdata, aes(x=Year, y=fed_diff, by=State)) + geom_line(aes(col=State))
ggplotly(p3 + eralines + zeroline + eralabels + plotlabs)

p3 + eralines + eralabels + zeroline + plotlabs + theme(legend.position = "none")

```

How much is the minimum wage worth in 2018 dollars? 
```{r}
eralines <- geom_vline(xintercept = newdata$xmin, alpha=0.25)
eralabels <- geom_text(data=newdata, x=newdata$xmin, y=15, label=newdata$fed.High.Value, angle=90, size=3, hjust=0.3)
fedline <- geom_line(data = (newdata %>% filter(State=='Federal (FLSA)')), col="black", size=1.2)
plotlabs <- labs(title = "State Minimum Wages in 2018 Dollars",
                y = '($)')
p4 <- ggplot(newdata, aes(x=Year, y=High.2018, by=State)) + geom_line(aes(col=State))

ggplotly(p4 + eralines + fedline + eralabels + plotlabs)

p4 + eralines + eralabels + plotlabs + fedline + theme(legend.position = "none")



```

```{r}
p5 <- ggplot(newdata, aes(x=Year, y=CPI.Average/High.2018, by=State)) + geom_line(aes(col=State))
eralines <- geom_vline(xintercept = newdata$xmin, alpha=0.25)
eralabels <- geom_text(data=newdata, x=newdata$xmin, y=15, label=newdata$fed.High.Value, angle=90, size=3, hjust=0.3)
fedline <- geom_line(data = (newdata %>% filter(State=='Federal (FLSA)')), col="black", size=1.2)
plotlabs <- labs(title = "Hours of min. wage work needed to meet CPI",
                y = 'Hours')
ggplotly(p5 + eralines + fedline + eralabels + plotlabs)

p5 + eralines + eralabels + plotlabs + theme(legend.position = "none")
```

Show a graphical summary of minimum wage growth or decline to keep up with inflation 
Federal, Iowa, California, DC, etc.
```{r}
p6.2018 <- ggplot(newdata %>% filter(State=='Federal (FLSA)' | State=='Iowa' | State =='California') %>% filter(Year >1990),
             aes(x=Year, y=High.2018, by=State)) + geom_line(aes(col=State)) + geom_point(aes(col=State))
eralines <- geom_vline(xintercept = newdata$xmin, alpha=0.25)
p6.real <- geom_line(data = newdata %>% filter(State=='Iowa' | State =='Federal (FLSA)' | State=='California')  %>% filter(Year>1990), aes(x=Year, y=High.Value, col=State), linetype="dashed")
p6.real.points <- geom_point(data = newdata %>% filter(State=='Iowa' | State =='Federal (FLSA)' | State=='California')  %>% filter(Year>1990), aes(x=Year, y=High.Value, col=State))
plotlabs <- labs(title = "A Look at Inflation",
                y = 'Wage')

ggplotly(p6.2018 + p6.real + plotlabs)
p6.2018 + p6.real + p6.real.points + plotlabs 
```

