---
title: "project_report"
author: "Allegra Frickel"
date: "4/25/2019"
output: html_document
---

---
title: "Project_report"
author: "Allegra Frickel"
date: "4/25/2019"
output: html_document
---

#Minimum Wages#

**Description**
Sourced from the United States Deptarment of Labor (DOL), this data set surrounds minimum wages in the United States from 1968 to 2017. Each wage provided in the data has a 2018 equivalency value.

**Columns**
- Year
- State
- Table Data: uncleaned data from the DOL
- Footnote: supplementary information from the DOL
- High.Value: Proposed minimum wage for those protected under the state's minimum wage laws
- Low.Value: Implemented minimum wage for those protected under the state's minimum wage laws
- High.2018: The High.Value in a 2018 dollar value
- Low.2018: The Low.Value in a 2018 dollar value
- CPI.Average: Average Consumer Price Index for the corresponding year. Out of a basket of goods accessible to everyone, CPI average is the average number of goods purchased by one consumer over the course of the year. CPI stays the same, but the price changes over the years. Thus, CPI average is used to calculcate a year's 2018-equivalent value.

**Questions Raised**

1. Which states showed the highest growth and which states showed the lowest growth of minimum wage?
2. Compare the minimum wage to the Consumer Price Index (CPI)
3. Explore if the gap between minimum wage and CPI has narrowed between the 2018-equivalent highs and lows as the years increase.
4. Show a graphical summary of minimum wage's growth or decline to keep up with inflation.

**Pipeline for Cleaning**

The data was already cleaned on Kaggle, so the cleaning is very minimal. Outside of deleting the two columns that did not contain valuable information, the NA values, data types, factoring, etc. was already done. It is logical to delete Table_Data and Footnote columns, these are just notes and not useful. For graphing, the data will be divided into pre-1990 and 1990-2017.

```{r}
library(tidyverse)
knitr::opts_chunk$set(fig.width = 15, fig.height=10) # change figure size in Rmd/HTML
# Load data from CSV
data <- read.csv('MinimumWageData.csv')
names(data)
head(data, 2)
```

```{r}
drops <- c("Table_Data", "Footnote")
data <- data[, !(names(data) %in% drops)]
head(data, 3)
```

```{r}
pre1990 <- filter(data, Year < 1990)
post1990 <- filter(data, Year >= 1990)
```

**Exploratory Analysis**

**Visualization**

**Answers to Questions**

**Conclusion**
include limitations here, what we would do differently and explore more